Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}

### 2018-06-28 paper.tex
add another bunch of todo's and fix other stuff

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{pgfplots}
\usepackage{listings}
\usepackage{multirow}
\usepgfplotslibrary{statistics}
\usepackage{dblfloatfix} %enable fig at bottom of page
%\input{macros.tex}

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. \todo{we should not use this term} WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.

\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters. The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters). If no order is specified, the VM assumes the byte ordering (ASCII order) is the order to compare against, and compares directly the bytes of the ByteStrings. If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order. Having the order optional allows to have quicker string comparison in the common case where the ASCII order is used. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

In the paper we will focus on the comparison of two strings with the default ASCII ordering and measure the performance only of this case. The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non ASCII are considered uncommon.

@@

\paragraph{\legBase.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level\footnote{The bytecode compiler moves the upper limit of the loop ahead of the loop, so (length1 min: length2) is not compute at each iteration.}. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.

@@

%Vspce to fit all on page.
\vspace{-0.1cm}
	\paragraph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). 
\vspace{-0.1cm}
	\paragraph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
\vspace{-0.1cm}
	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
\vspace{-0.1cm}
	\paragraph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we count all the Slang implementations as well as the Cog's RTL one.
\vspace{-0.1cm}
	\paragraph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
\vspace{-0.1cm}
	\paragraph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
\vspace{-0.1cm}
	\paragraph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\vspace{0.2cm}

@@

\subsection{Methodology}

>>>>

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{pgfplots}
\usepackage{listings}
\usepackage{multirow}
\usepackage{soul}
\usepgfplotslibrary{statistics}
\usepackage{dblfloatfix} %enable fig at bottom of page
%\input{macros.tex}

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. \todo{we should not use this term. Maybre prefer Latin-1 standard} WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (\st{ASCII or }Unicode) or using different orders, for example, the case insensitive order.

\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters. The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters). If no order is specified, the VM assumes the byte ordering (\st{ASCII} \sk{Unicode} order) is the order to compare against, and compares directly the bytes of the ByteStrings. If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order. Having the order optional allows to have quicker string comparison in the common case where the \st{ASCII} \sk{Unicode} order is used. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

In the paper we will focus on the comparison of two strings with the default \st{ASCII}\sk{Unicode} ordering and measure the performance only of this case. The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non- \sk{Unicode }ASCII are considered uncommon.

@@

\paragraph{\legBase.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level\footnote{The bytecode compiler moves the upper limit of the loop ahead of the loop, so (length1 min: length2) is not compute at each iteration.}. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII \sk{Unicode} ordered comparison, we show the code only of baselineCompareWith:.

@@

%Vspce to fit all on page.
\vspace{-0.1cm}
	\paragraph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). 
\vspace{-0.1cm}
	\paragraph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
\vspace{-0.1cm}
	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, \todo {} \st{which arguably can be used to evaluate the complexity level of each implementation.}
\vspace{-0.1cm}
	\paragraph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we count all the Slang implementations as well as the Cog's RTL one.
\vspace{-0.1cm}
	\paragraph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
\vspace{-0.1cm}
	\paragraph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
\vspace{-0.1cm}
	\paragraph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\vspace{0.2cm}

@@

\subsection{Methodology}
\todo{so now I have to figure out how to actually bootstrap errors computed that way}
